---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# datavalr: Tools for Data Validation and Quality Assessment

## Why should I use this package? 

Data quality is critical for accurate analysis and decision-making. The **datavalr** package provides an essential toolkit for validating datasets, diagnosing common data issues, and performing quality control checks. Whether you are working with real-world messy data or aiming to ensure data integrity in predictive modeling, **datavalr** offers robust and intuitive functions for exploratory data analysis and cleaning.

The goal of datavalr is to diagnose primarily data entry errors, flagging them for collaborators to fix before further statistical analysis. The package can denote outliers, misclassification of data frame columns, showcase missingness in data frames, and diagnose other data entry errors. 

### Key Features
- **Diagnose Empty Values**: Quickly identify columns with empty strings, assessing their prevalence and offering a data frame with NA's replacing the empty strings. 
- **Summarize Columns**: Generate insights about column data types and missingness
- **Track Outliers**: Detect and visualize possible mice outliers by flagging irregular weight trends over time. 
- **Validate data consistency**: Ensure chronological order and logical consistency of time series mice data. 
- **Audit Betting Metrics**: Evaluate and flag anomalies in betting metrics like spreads and over/under lines.

Designed with ease of use in mind, **datavalr** streamlines data validation and enhances your workflow.

## How Do I Access the Package? 

You can install the development version of datavalr from [GitHub](https://github.com/oscar-mercado/datavalr) with:

``` r
# install.packages("pak")
pak::pak("oscar-mercado/datavalr")
```

Otherwise install the `devtools` library and run the following code:

```r
devtools::install_github("oscar-mercado/datavalr")
```

## How Do I Use the Package?

Here are some examples to demonstrate the core functionality of **datavalr**: 

```{r example}
library(datavalr)
df <- footballbetting23
summarize_columns(df)
```

The first example above loads the `footballbetting23` dataset in the library and uses the `summarize_columns()` function. This generates a comprehensive data table summarizing the columns of our dataframe including data types and missing values for each column. This helps collaborators determine data entry errors if they want columns to have different classes or further analyze the missingness in the data. Through further exploration of the football data frame, I observe something unique: the `weather_detail` column contains empty strings as possible values but are not counted as missing

```{r pressure, echo = FALSE}
unique(df$weather_detail)
```

In this case, we can use the `diagnose_empty_values()` function to determine the count of empty string values and replace them with NA's if collaborators desire: 

```{r}
diagnosis <- diagnose_empty_values(df)
head(diagnosis$Modified_Data)
diagnosis$Empty_Values_Summary
```

## Additional Information

For more advanced examples and explanations, please refer to the package vignette: 

```r
vignette("datavalr")
```

## Contributions

Contributions are welcome! Feel free to fork the repository, submit issues, or propose enhancements. Contact me at oscarmercado@college.harvard.edu for any inquiries.

## License

This package is licensed under GPL (>= 3). See the LICENSE file for more details.
